Two applications of logic programming to Coq

### 2020-11-02 papers/types2020/elab.tex
Finished a draft of section 5.

\newpage
\section{Elaboration of external proof certificates for the Coq
  kernel}

@@

There are many theorem provers for intuitionistic logic for which a
successful proof is not the kind of detailed $\lambda$-term required
by the Coq kernel.  Most prover provide, instead, a trace of some key
aspects of a proof: some details are often not included in such a
trace.  For example:
\begin{enumerate}

@@

As we have describe above, the Foundational Proof Certificate (FPC)
framework has been designed for formally defining a rich collection of
proof certificates that do not necessarily contain full details.
Proof checkers for FPC certificates can be written using simple logic
programs: the standard implementation mechanisms of unification and
backtracking search can then be used to reconstruct missing proof
details \cite{miller17fac}.  Such proof checking kernels can also be
used to elaborate such certificates into fully detailed proofs
\cite{blanco17phd,blanco17cade}.

We describe here briefly how to use the technology behind FPCs and
logic programming in order to provide a flexible approach to
connecting external provers of intuitionistic logic to Coq.
%
Following the general outline that has been described in
\cite{blanco17cade}, we assume that the following step are taken.
%
\begin{enumerate}
\item Have an external prover output some form of proof evidence
  (proof certificate) for formulas it claims are theorems.

@@

To illustrate the possibilities ***

** We limit our attention here to first-order intuitionistic logic:
that is, we do not consider here inductive definitions.

** Note that with intuitionistic logic, there is much less ambiguity
of polarities possible.  In particular, only the conjunction and the
atomic formulas can have ambiguous polarizations.  The two
conjunctions are familiar to those working with typed
$\lambda$-calculus:

the negative conjunction (corresponding to $\with$ in linear logic)
corresponds to surjective pairing.

the positive conjunction (corresponding to $\with$ in linear logic)
corresponds to ***.

The choice between negated atoms and positive atoms is similar to the
choice in representing $\beta$-normal terms using either head-normal
form or administrative normal form.

We built the fpccoq system to demonstrate the feasibility of using
logic programming to check proof certificates produced by external
provers and to elaborate such certificates into proofs with sufficient
details that they can be given directly to Coq's kernel to
check.  Note that one does not need to trust the external prover nor
the fpccoq tool-chain: one only needs to trust the Coq kernel.

De Bruijn
decide depth
simply typed lambda terms

\newpage

>>>>

%\newpage
\section{Elaboration of external proof certificates for the Coq
  kernel}

@@

There are many theorem provers for intuitionistic logic
\cite{raths07jar} for which a successful proof is not the kind of
detailed $\lambda$-term required by the Coq kernel.  Often, such
provers provide no information about the proofs they discover.  To the
extent that some prove evidence is output after a successful run, such
evidence is usually just a trace of some key aspects of a proof: some
details are often not included in such a trace.  For example:
\begin{enumerate}

@@

As we have describe in Section~\ref{sec:three}, the Foundational Proof
Certificate (FPC) framework has been designed for formally defining a
rich collection of proof certificates that do not necessarily contain
full details.  Proof checkers for FPC certificates can be written
using simple logic programs: the standard implementation mechanisms of
unification and backtracking search can then be used to reconstruct
missing proof details \cite{miller17fac}.  Such proof checking kernels
can also be used to elaborate such certificates into fully detailed
proofs \cite{blanco17phd,blanco17cade}.

We describe here briefly how to use the technology behind FPCs and
logic programming in order to provide a flexible approach to
connecting external provers of first-order intuitionistic logic to
Coq.
%
Following the general outline that has been described in
\cite{blanco17cade}, we assume that the following step are taken.
%
\begin{enumerate}
\item Have an external prover output some form of proof evidence
  (proof certificate) for formulas it claims are theorems.

@@

To illustrate this approach to elaborating externally provided proof
structures into terms the Coq kernel can check, we consider here
examples take from the file \verb+coq_fpc.v+, which is part of the
repository \url{https://github.com/proofcert/fpc-elpi/}.
Figure~\ref{fig:debruijn} presents the specific FPC that defines the
proof semantics of using the De Briujn notions for untyped
$\lambda$-terms as proof evidence of a proof in intuitionistic
propositional logic.  That FPC definition, along with a proof checker
capable of synthesizing Coq proof terms during proof checker, makes it
possible to use the term denoting the untyped $\lambda$-term
$\lambda x (x (\lambda y (\lambda z (x (\lambda u z)))))$

\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
                                                    [lambda (apply 1 [])])])]))
\end{lstlisting}
as a proof of following theorem of Coq.
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
Theorem dneg_peirce_mid : forall P Q: Prop, (((((P -> Q) -> P) -> P) -> Q) -> Q).
\end{lstlisting}

\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
\caption{The FPC definition of De Bruijn notation as proof evidence.}
\label{fig:debruijn}
\end{figure}

%\newpage
